---
title: "Hyperoxic Load & Sepsis"
author: "C.V. Cosgriff, MD/MPH Student, Harvard Chan School"
output:
  html_document: default
---

## Libraries & Database Connection
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(broom)
library(lubridate)
library(stringr)
library(RPostgreSQL)
library(MIMICbook) #JRaffa's MIMIC code book
library(epitools) # For calculating OR's
library(tableone) # Generating Table 1
library(sjstats) # Some useful stats tools e.g. hoslem_gof
library(sjPlot)

library(rstudioapi)

library(dslabs) # Dr. Irrizary's library 
ds_theme_set() # Dr. Irrizary's theme

# Database access; you'll need to change 10.8.0.1 to the server you are using
# and you will also need to run ths in RStudio as I use the RStudio API to ask
# for a password. Edit this definition for your environment as need.
drv <- dbDriver("PostgreSQL")
mimic <- dbConnect(drv, dbname = "mimic", host = "10.8.0.1", port = 5432,
                   user = "mimicuser", 
                   password = askForPassword("MIMIC Password: "))
```

## Cohort Extraction
The initial cohort table was began in sql with the following code (adapted from
Alistair Johnson and Tom Pollard's code). 

```{sql, eval = FALSE}
SET search_path = 'mimiciii';

DROP MATERIALIZED VIEW IF EXISTS hyperoxia_sepsis_baseCohort CASCADE;
CREATE MATERIALIZED VIEW hyperoxia_sepsis_baseCohort AS

WITH icuStaysBase AS 
(
SELECT icu.subject_id, icu.hadm_id, icu.icustay_id, adm.hospital_expire_flag
, EXTRACT(EPOCH FROM outtime - intime)/60.0/60.0/24.0 as icu_length_of_stay
, EXTRACT('epoch' from icu.intime - pat.dob) / 60.0 / 60.0 / 24.0 / 365.242 as age
, pat.gender, adm.ethnicity
, CASE
        WHEN adm.deathtime BETWEEN icu.intime and icu.outtime
            THEN 1
        -- sometimes there are typographical errors in the death date, so check before intime
        WHEN adm.deathtime <= icu.intime
            THEN 1
        WHEN adm.dischtime <= icu.outtime
            AND adm.discharge_location = 'DEAD/EXPIRED'
            THEN 1
        ELSE 0
        END AS icustay_expire_flag

FROM icustays icu
INNER JOIN patients pat /* Inner join because we only need patients who had an ICU stay */
  ON icu.subject_id = pat.subject_id
INNER JOIN admissions adm /* Inner join because we only care about admissions which have an ICU stay */
    ON icu.hadm_id = adm.hadm_id
)
, ventDurations AS
(
SELECT vd.icustay_id, vd.duration_hours
FROM ventdurations vd
)
, oasisscore AS
(
SELECT oa.icustay_id, oa.oasis
FROM oasis oa
)
, elixhauser AS
(
SELECT el.hadm_id,
    el.elixhauser_vanwalraven AS score
    FROM elixhauser_quan_score el
)
SELECT b.*, vd.duration_hours AS vent_dur, s.angus
, CASE
	WHEN b.icu_length_of_stay <= 0.1666667 THEN 1
    ELSE 0 END
    AS exclusion_los
, CASE
	WHEN b.age < 16 then 1
    ELSE 0 END
    AS exclude_age
, el.score AS elixhauser, oa.oasis
, CASE
	WHEN (SELECT count(vpd.icustay_id) FROM vasopressordurations vpd WHERE b.icustay_id = vpd.icustay_id AND vpd.duration_hours > 0) > 0 THEN 1
    ELSE 0 END
    AS vasopressor_flag
FROM icuStaysBase b
INNER JOIN oasis oa on b.icustay_id = oa.icustay_id
INNER JOIN elixhauser el on b.hadm_id = el.hadm_id
INNER JOIN ventDurations vd ON b.icustay_id = vd.icustay_id /* Our cohort it will only include people who were ventilated */
INNER JOIN angus_sepsis s on b.hadm_id = s.hadm_id;
```

Next, the base cohort can be pulled from the database.
```{r}
cohort <- dbGetQuery(mimic, 
                        "SELECT * FROM mimiciii.hyperoxia_sepsis_baseCohort")
```

This cohort contains all of the ICU stays for which ventilation data existed
and each row actually corresponds to a ventilation duration recording. 
We will now build the subject cohort that will eventually be used for analysis
and as we go we will tidy the data.

Before we begin inclusion/exclusion, lets do some very basic cleaning.

There are ages of 300 in the dataset because of HIPPA, so we set them to the 
median of that group per MIT LCP.

```{r}
cohort$age[cohort$age>100] <- 91.4
```

Lets also recode ethnicity since there are too many groups based on the MIMIC
ethnicities.
```{r}
ethnicities <- c("WHITE", "BLACK", "HISPANIC", "ASIAN")

for (i in 1:length(ethnicities)) {
  cohort$ethnicity[str_detect(cohort$ethnicity, ethnicities[i])] <- ethnicities[i]
}
cohort$ethnicity[!(cohort$ethnicity %in% ethnicities)] <- "OTHER"
cohort$ethnicity <- as.factor(cohort$ethnicity)
cohort$ethnicity <- relevel(cohort$ethnicity, ref = "WHITE")
```

Now we can begin including/excluding patients to form the study cohort. We start
by introducing a function for examining the number of unique subjects in our
study.

```{r}
subjectCount <- function(cohort) {
  cohort %>% distinct(subject_id) %>%
    summarise(count = n())
}

subjectCount(cohort)
```

We are starting with 23,256 patients. We have to be careful about inclusion and
exclusion because some flags apply to hospital admissions, and others apply to
ICU stays specifically; in the end we want to exclude any subject who has met
exclusion criteria at some point. We also have elected to not deal with 
integrating data across multiple admissions and so we'll first remove subjects
with multiple admissions. This step has to be done first because if we exclude 
an admission because, for example, sepsis didn't occur during it, we may 
inadverdently get rid of the 'second' admission; we'll have kept a patient with
more than one admission by accident.

```{r}
adm_count <- cohort %>% select(subject_id, hadm_id) %>% 
  distinct(hadm_id, .keep_all = TRUE) %>% group_by(subject_id) %>% 
  summarise (count = n()) %>% 
  arrange(-count)
adm_count <- adm_count %>% filter(count == 1)
cohort <- cohort %>% filter(subject_id %in% adm_count$subject_id)
subjectCount(cohort)
```
Next we'll exclude patients with multiple ICU stays. Although we don't want to
throw away data, this is useful for the following reason: consider a patient
goes to the ICU, gets blasted with O2, and makes it to the floor, and then 
returns to the ICU the next day and dies. Since we are excluding patients who 
die in the ICU we will throw away this initial survival. Because all of this
challenges analysis (lots of competing risk) for this work we will focus on
single admission, single ICU stay patients.

```{r}
icu_count <- cohort %>% select(subject_id, icustay_id) %>% 
  distinct(icustay_id, .keep_all = TRUE) %>% group_by(subject_id) %>% 
  summarise(count = n()) %>% arrange(-count)
icu_count <- icu_count %>% filter(count == 1)
cohort <- cohort %>% filter(subject_id %in% icu_count$subject_id)
subjectCount(cohort)
```

Now we have a data frame composed of all of the ventilation instances for patients
with 1 hadm_id and 1 icustay_id. Other covariates shouldn't vary with respect
to ventilation duration, and so we can now simply collapse the ventilation 
durations into single observations. If we do this right, the subject count
shouldn't change. For things that we know are constant by subject_id at this 
point (e.g. ICU length of stay) we can just take the mean in the summarise 
function.

```{r}
cohort <- cohort %>% group_by(subject_id, hadm_id, icustay_id, gender, 
                                    ethnicity, first_careunit) %>%
  summarise(age = mean(age), icu_los = mean(icu_length_of_stay), 
            vent_duration = sum(vent_dur), angus = mean(angus), 
            elixhauser = mean(elixhauser), oasis = max(oasis),
            exclusion_icu_expire = any(icustay_expire_flag == 1),
            exclusion_age = any(exclude_age == 1), exclusion_los = any(exclusion_los),
            vasopressor_flag = any(vasopressor_flag == 1), 
            hospital_expire = any(hospital_expire_flag == 1)) %>% ungroup() %>%
  select(subject_id, icustay_id, age, gender, ethnicity, first_careunit, elixhauser, oasis, icu_los, vent_duration, 
         angus, vasopressor_flag, hospital_expire, exclusion_icu_expire, exclusion_age, 
         exclusion_los)
subjectCount(cohort)
```

Now, with each row corresponding to a potential subject, we can begin to exclude
patients.

We start by keeping only septic patients by the Angus criteria.

```{r}
cohort <- cohort %>% filter(angus == 1)
subjectCount(cohort)
```

We then exclude patients below age 16, those who died before completing their
ICU stay, and those who had an LOS <4h. These were encoded as flags during the 
SQL pull and so we can exclude them same as we did with sepsis above.

```{r}
cohort <- cohort %>% filter(exclusion_age == 0)
subjectCount(cohort)
cohort <- cohort %>% filter(exclusion_icu_expire == 0)
subjectCount(cohort)
cohort <- cohort %>% filter(exclusion_los == 0)
subjectCount(cohort)
cohort <- cohort %>% filter(vent_duration > 4)
subjectCount(cohort)
```

This leaves us with 4,585 subjects. We now have our baseline cohort, although
we have not taken into account any desaturations. This however cannot be done
until we have developed a profile of their oxygenation exposures. 

Before continuing let us just check if there is any missig data in our current
dataset.

```{r}
sapply(cohort, function(x) sum(is.na(x)))
```

None, great! On to exposures.


## Exposure Data

The goal is to generate a table containing all of the necessary SpO2 data. This
is computationally challenging because not only are their millions of SpO2 data,
but we also only want to only grab SpO2 data that falls within our patients
ventilation times.

We begin by using the following code to generate a table containing all of our 
SpO2 data.

```{sql eval=FALSE}
SET search_path = 'mimiciii';
DROP MATERIALIZED VIEW IF EXISTS spO2 CASCADE;
CREATE MATERIALIZED VIEW spO2 AS
SELECT ce.row_id,
    ce.subject_id,
    ce.hadm_id,
    ce.icustay_id,
    ce.itemid,
    ce.charttime,
    ce.storetime,
    ce.cgid,
    ce.value,
    ce.valuenum,
    ce.valueuom,
    ce.warning,
    ce.error,
    ce.resultstatus,
    ce.stopped
FROM mimiciii.chartevents ce
WHERE ce.itemid = 646 OR ce.itemid = 220277;
```

An attempt was then made to run the following query.

```{sql eval = FALSE}
SET search_path = 'mimiciii';
DROP MATERIALIZED VIEW IF EXISTS spO2_vent CASCADE;
CREATE MATERIALIZED VIEW spO2_vent AS
SELECT ce.row_id,
    ce.subject_id,
    ce.hadm_id,
    ce.icustay_id,
    ce.itemid,
    ce.charttime,
    vd.starttime,
    vd.endtime,
    ce.storetime,
    ce.cgid,
    ce.value,
    ce.valuenum,
    ce.valueuom,
    ce.warning,
    ce.error,
    ce.resultstatus,
    ce.stopped
FROM mimiciii.spO2 ce
INNER JOIN ventdurations vd ON ce.icustay_id = vd.icustay_id
WHERE vd.starttime <= ce.charttime AND vd.endtime >= ce.charttime;
```

However, this query would not run on my server (which has numerous VMs) because
of a lack of memory. To handle this I elected to collect all of the SpO2 data
into a vector from which I could then filter it down to the current cohort.

An attempt was made to pulled the necessary data right into a vector, but this
also failed (as such, eval was set to false for knitr). This was not due to
memory, but because my server is remote and the transfer was too large for it
to be usable.

```{r eval = FALSE}
spO2 <- dbGetQuery(mimic, 
                        "SELECT * FROM mimiciii.spO2")
```

The full SpO2 table was exported and gziped. It was then stored locally in 
./data which is not in the repo. If you wish to recreate this project you'll
need to run the SQL code above (the one to generate mimiciii.spO2) and then
export that file, gzip it, and store it in "./data/" where it can be accessed
locally.


```{r, warning=FALSE, message=FALSE}
spO2_columns <- c("row_id", "subject_id", "hadm_id", "icustay_id", "item_id", 
              "charttime", "storetime", "cgid", 
              "value", "valuenum", "valueom", "warning", "error", "resultstatus",
              "stopped")
spO2 <- as.data.frame(read_csv("./data/spO2.csv.gz", col_names = spO2_columns))
```

And now we can filter out everyone from spO2 who isn't in our cohort.

```{r}
spO2 <- spO2 %>% filter(spO2$subject_id %in% cohort$subject_id) %>%
  select(subject_id, icustay_id, charttime, valuenum)
```

Now to get this data such that we only have SpO2 data relating to vents, we will
begin by reloading the LCP's vent duration table.

```{r eval = FALSE}
vent <- dbGetQuery(mimic, "SELECT * FROM mimiciii.ventdurations") %>%
  filter(icustay_id %in% cohort$icustay_id) %>% 
  select(icustay_id, starttime, endtime)
```

However, it was at this time that I discovered a bug in RPostgresSQL which causes
it to throw away the time associated with a POSIX datetime. There are discussions
about it on stack overflow, but needless to say it can be solved by directly 
downloading the CSV and loading it locally, and so once again this is what was
done.

```{r, message=FALSE, warning=FALSE}
#vent <- as.data.frame(fread(paste0("\"C:\\Program Files\\Git\\bin\\sh.exe\" zcat < ", "./data/ventdurations.csv.gz"), showProgress = FALSE)) 
vent_columns <- c("icustay_id", "ventnum", "starttime", "endtime", "duration_hours")
vent <- as.data.frame(read_csv("./data/ventdurations.csv.gz", col_names = vent_columns))
vent <- vent %>%
  filter(icustay_id %in% cohort$icustay_id) %>% 
  select(icustay_id, starttime, endtime)
```


Before joining, lets just make sure we have our data columns as dates.

```{r}
vent$starttime <- as_datetime(vent$starttime, tz = "EST")
vent$endtime <- as_datetime(vent$endtime, tz = "EST")
spO2$charttime <- as_datetime(spO2$charttime, tz = "EST")
```

And now we perform a join of these spO2 data with the vent data; with that
we can then filter the table to only include spO2 recordings from within the
patients ventilation period by using the spO2 charttime and the start and end
time of ventilation.

```{r}
spO2_vent <- inner_join(spO2, vent, by = c("icustay_id"))
spO2_vent <- spO2_vent %>% filter(charttime >= starttime & charttime <= endtime)
```

With that we now have all of our SpO2 data. Now we'll need to summarise it in a
way that it becomes an exposure. Before we can do that though we need to check it
for missing values.

```{r}
sapply(spO2_vent, function(x) sum(is.na(x)))
```

We are missing a meager 323 single recordings of 897,879, and so these can be
disregarded. 

```{r}
spO2_vent <- spO2_vent %>% na.omit()
```

## Defining Hyperoxia from SpO2 Data

With this spO2 data in hand we need to now develop a summary measure, and then
create an exposure variable. There are some important considertions in doing in
approaching this. First off, we have to be careful about capturing distinct 
ventilation periods. If we write a function that sums across the patient's values
in the ICU the gap between two vent periods may be captured as part of the
summary value. We also need to consider how to treat outliers and impossible 
values. We can begin by looking at our spO2 values.

```{r}
summary(spO2_vent)
```

Looking at the values for the spO2 data we see that the minima is 0 and the 
maxima is 1000. We should assess how many values fall in this impossible range.

```{r}
summary(spO2_vent %>% filter(valuenum > 100))
length(spO2_vent$valuenum[spO2_vent$valuenum > 100])
```
So there are only 3 values above 3. With respect to the other side of the 
distribution, how many recordings are below 50.

```{r}
length(spO2_vent$valuenum[spO2_vent$valuenum < 50])
```

187 values below 50 out of 897,556 is reassuring as they are unlikely to be
very important in challenging the valdity of our results. Even fewer are below 10
and some of these many thousands of recordings are likely just entry errors
(e.g. 10 was just 100 with a dropped 0). For now, we'll just exclude the three
records above 100 as a value of 1000 could damage the summarization, but the 
very low values can be kept as they are unlikely to cause much instability. 

```{r}
spO2_vent <- spO2_vent %>% filter(valuenum <= 100)
```

Before summarizing we can look at the oxygen saturation trajectories for some of
the subjects to get a better understanding of how the data looks.

```{r}
sample_id <-  sample(cohort$subject_id, 1) # 59970 a good example of two vent periods
spO2_sample <- spO2_vent %>% filter(subject_id == sample_id)
spO2_sample <- spO2_sample[order(spO2_sample$charttime),]
spO2_sample %>% ggplot(aes(x = charttime, y = valuenum)) + geom_point() +
  geom_smooth(se = F)
```

Running the above code a few times demonstrates various patterns of SpO2 
recordings over a hospital stay. 


Now lets examine the trajectories if we normalize the data such that 98 is taken
as a reference point for normalcy; we can plot these as a line and use geom_area
to get an idea of the AUC. 

```{r}
sample_id <-  sample(cohort$subject_id, 1) # 59970 a good example of two vent periods
spO2_sample <- spO2_vent %>% filter(subject_id == sample_id)
spO2_sample <- spO2_sample[order(spO2_sample$charttime),]
spO2_sample$valuenum <- spO2_sample$valuenum - 98
spO2_sample %>% ggplot(aes(x = charttime, y = valuenum)) + geom_point() +
  geom_line() + geom_area(aes(fill = valuenum > 0), alpha = 0.5)
```

Building off the ideas expressed by Du and Shen, the goal is to develop a 
measure of the "hyperoxic load" by calculating the AUC of an SpO2 per time
curve. By taking all values relative to 98 a patient's "load" benefits from 
periods below 98, and suffers from periods above. The cutoff of 98% can be
addressed in a sensitivity analysis. Of note, we plan to exclude patients
who have had more than 2 hypoxic events, and, as such, we will idealy have a 
majority of patients who can have as much as +2 and as low as -4, and then 
we'll integrate these curves with respect to time.

To get a load per patient, we'll write a function that takes chart times
and recordings, orders them, and then generates a curve. We'll take all values
relative to 98, and then we'll 












