---
title: 'Hyperoxia & Sepsis: Preliminary Report'
author: "C. V. Cosgriff, MD/MPH Student, Harvard Chan School"
output:
  pdf_document: default
---

## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(stringr)
library(RPostgreSQL)
library(MIMICbook)
library(epitools)
library(tableone)
library(sjstats)
library(sjPlot)
library(broom)
library(rms)
library(quantreg)
library(pROC)
library(separationplot)
library(heatmapFit)
library(rstudioapi)
```

## Data Loading 
```{r message = FALSE, warning = FALSE}
drv <- dbDriver("PostgreSQL")
mimic <- dbConnect(drv, dbname = "mimic", host = "10.8.0.1", port = 5432,
                   user = "mimicuser", 
                   password = askForPassword("MIMIC Password: "))
```




## Load the Data
The dataset was generated in multiple steps; these are found  cohort3.sql. 
Because of a computational limitation, we couldn't pull directly from Postgres 
for the SpO2 items.
```{r, warning = FALSE, message = FALSE}
cohort.df <- dbGetQuery(mimic, 
                        "SELECT * FROM mimiciii.hyperoxia_sepsis_covariates2")

colnames.spo2 <- c("row_id", "subject_id", "hadm_id", "icustay_id",
                   "charttime", "ventstart", "ventend", "valuenum")
spo2.df <- read_csv("D:/HST953_data/hyperoxia_sepsis_spo2.csv", col_names = colnames.spo2)
```


## Cleaning the Data: Part 1
Lets screen for missing data in general.
```{r}
sapply(cohort.df, function(x) sum(is.na(x)))
```

No missing data in the cohort.

```{r}
sapply(spo2.df, function(x) sum(is.na(x)))
```

Missing 70 values out of 374578; trivial to exlcude.

```{r}
spo2.df <- spo2.df %>% na.omit()
```

Now lets just check if we have any nonsensical values for Spo2.
```{r}
spo2.df[spo2.df$valuenum > 100,]$valuenum
length(spo2.df[spo2.df$valuenum > 100,]$valuenum)
```
Only two out of 374508; also trivial to exclude.

What about <94? Because we excluded patients with>2 SpO2 readings below 94, we 
made the implict assumption that the two or less they had were recording errors.
As such, these should be excluded, but lets check how much data that would be 
exluding.

```{r}
length(spo2.df[spo2.df$valuenum < 94,]$valuenum)
```


Here we have 2412 records of 374508; these were presumed errors by study design,
and in this case account for a very small proportion of the data. The approach
used to determine hyperoxic time won't include them to begin with, and so they
can be excluded. The very small number of them indicates that the extraction 
worked properly, and that they represent an exlcudable portion of our dataset.
```{r}
spo2.tidy.df <- spo2.df %>% filter(valuenum <= 100 & valuenum >= 94)
```

We want to also check if there any vent stretches for which only one SpO2 recording
occurred; this should be rare considering we exclude anyone with <4h on the vent
because we assume that there is at least one recording per hour.
## 
```{r}
spo2.count.df <- spo2.tidy.df %>% group_by(icustay_id) %>% 
  summarise(count = n()) %>% filter(count < 4)
print(spo2.count.df)
```

There was only one ICU stay with this problem, and this was the only ICU stay
for that patient, and so we'll remove it; as it is only one observation this is
unlikely to induce any bias, and it would likely be more biased to include a 
pateint who was on a vent for >=4 hours but had only 1 SpO2 recording.

```{r}
cohort.df <- cohort.df %>% filter(!(cohort.df$icustay_id %in% spo2.count.df$icustay_id))
spo2.tidy.df <- spo2.tidy.df %>% filter(!(spo2.tidy.df$icustay_id %in% spo2.count.df$icustay_id))
```

MIMIC-III codes age > 89 as ~300 to protect confientiality, and the manual says
the median age for that group was 91.4. We'll also code gender M = 1, F = 0.
```{r}
cohort.df[cohort.df$age > 89, "age"] <- 91.4
cohort.df$gender <- as.numeric(cohort.df$gender == "M")
```

We've added ethnicity to the dataset (11/11/17), but it is a factor with many
possibilities. For these we will pool into white, black, asian, and other.

```{r}
ethnicities <- c("WHITE", "BLACK", "HISPANIC", "ASIAN")

for (i in 1:length(ethnicities)) {
  cohort.df$ethnicity[str_detect(cohort.df$ethnicity, ethnicities[i])] <- ethnicities[i]
}
cohort.df$ethnicity[!(cohort.df$ethnicity %in% ethnicities)] <- "OTHER"
```


Before we cleanup the data further, we need to get a tidy data frame, and so
we need to take all of SpO2 data, calculate hyperoxic times, and then join this
information to our cohort table.

## Hyperoxic Durations

### Duration of Exposure
The following function was developed to calculate the amount of time hyperoxic.
Intervals are calculated from chart times, and SpO2 values are considered 
constant between each interval.

```{r}
hyperoxicTime <- function(times, value) {
  icustay.spO2 <- data.frame(time = times, value = value)
  icustay.spO2 <- icustay.spO2[order(icustay.spO2$time),]
  intervals <- as.numeric(icustay.spO2$time - lag(icustay.spO2$time))
  intervals <- lead(intervals)
  intervals[is.na(intervals)] <- 0
  
  # Intervals are usually ~60, we will tolerate as many as three missed entries
  # or other errors; beyond that the result will be tossed.
  intervals[intervals >= 3*60] <- 0
  
  icustay.spO2 <- icustay.spO2 %>% mutate(duration = intervals)
  o2levels <- icustay.spO2 %>% group_by(value) %>% 
    summarise(time_at = as.numeric(sum(duration))) %>% arrange(desc(value))
  o2levels[is.na(o2levels)] <- 0
  colnames(o2levels)[1] <- "satValue"
  dur.100 <- as.numeric(o2levels[o2levels$satValue == 100, 2])
  dur.99 <-  as.numeric(o2levels[o2levels$satValue == 99, 2])
  
  # Because some patients will never have any values at 99 or 100, we need to
  # return a 0 for these times; however they will be null. We need to clean 
  # this up.
  if (is.null(dur.100) | length(dur.100) <= 0) {
    dur.100 <- 0
  }
    if (is.null(dur.99) | length(dur.99) <= 0) {
    dur.99 <- 0
  }
  if (is.na(dur.100)) {
    dur.100 <- 0
  }
  if (is.na(dur.99)) {
    dur.99 <- 0
  }
  
  # Now we can calculate the hyperoxic time and return it
  hyperoxic.time <- (dur.100 + dur.99) / 60
  return(hyperoxic.time)
}
```

With this function, durations of hyperoxia (>98%) were calcualted for each ICU 
stay ID.
```{r}
spo2.profiled <- spo2.tidy.df %>% 
  group_by(icustay_id) %>% summarise(hyperoxic_dur = as.numeric(hyperoxicTime(charttime, valuenum)))
```

Now we can data merged with our cohort data on icustay_id.
```{r}
hs.df <- inner_join(cohort.df, spo2.profiled, by = "icustay_id")
```


## Cleaning Part 2

There are ~290 patients in the cohort who have multiple hospital admissions. 
This is a problem: consider a patient who comes into the hospital in February,
is discharged, and then returns in March, and is discharged again. Consider this
patient dies in June. Now consider that they had a significant amount of the
exposure in the first stay but not the second. They would therefore be in both
arms of the study and produce only 1 outcome. Instead, we could look at first
admission or throw these patients out. We'll start with throwing them out,
and store a frame with earliest for sensitivity analysis later.


```{r}
# To remove, we need a list of subject IDs of any patient with more than one
# admission.
check <- hs.df %>% group_by(subject_id) %>% summarise(count = n())
check <- check %>% filter(count < 2)
hs.df <- hs.df[hs.df$subject_id %in% check$subject_id,]

# This was a way of just including the earliest; we'll look at this in 
# sensitivity analysis.
#hs.firstAdmit.df <- hs.df %>% group_by(subject_id) %>% 
#  filter(admittime == min(admittime))

```

We have multiple ICU stays per hospital admission but eventually just want to 
examine a subjects one-year mortality following their discharge; so we'll summarize
their ICU stays into one admission with the following logic:
Because the OASIS score is calculated for each ICU stay, we'll take
use their highest OASIS as the OASIS for the overall admission. Also note, we
take the average of age, but it should be virtually identical across the admission.
Elixhauser doesn't change by ICU stay, and we assign pressors based on if it
they were given during any stay.
```{r}
hs.df <- hs.df %>% group_by(subject_id, hadm_id, admittime, gender, ethnicity, elixhauser, first_careunit) %>%
  summarise(oasis = max(oasis), age = mean(age), vasopressor_flag = as.numeric(any(vasopressor_flag == 1)),
            icu_los = sum(icu_length_of_stay), vent_dur = sum(vent_duration), 
            hyperoxic_dur = sum(hyperoxic_dur), hospital_expire = as.numeric(any(hospital_expire_flag == 1)),
            oneyear_expire = as.numeric(any(one_year_expire == 1)))
```

Our extraction contains data on hospital expiration in the event we want to examine
this as a secondary outcome. For now though we will proceed by removing any 
patient who died in the hospital. So our cohort is patients who survive their
hospitalization, and we are curious about one year mortality.
```{r}
#hs.df <- hs.df %>% filter(hospital_expire == 0) %>% ungroup()
hs.df <- hs.df  %>% ungroup()
```

Our vent and hyperoxic durations are in hours, but the extracted LOS is in days;
will flip it to hours for ease of comparison.

```{r}
hs.df$icu_los <- hs.df$icu_los * 24
```


Lastly, Leo wanted to examine hyperoxia as a categorical variable. To do this,
we'll generate groups based on quartiles with the bottom quartile as group 1 and
the top quartile as group 4. We'll do the same for ventilation in case we are 
curious about it later.
```{r}
groups <- 4
hs.df$hyperoxic_quartile <- 0
hs.df$hyperoxic_group <- 0

hs.df$hyperoxic_quartile <- with(hs.df, cut(hyperoxic_dur, 
                                breaks = quantile(hyperoxic_dur, 
                                                  probs = seq(0,1, by = (1/groups)), 
                                                  na.rm=TRUE), 
                                include.lowest = TRUE))

hyperoxia.levels <- levels(hs.df$hyperoxic_quartile)

for (i in 1:groups) {
  hs.df$hyperoxic_group[hs.df$hyperoxic_quartile == hyperoxia.levels[i]] <- i
}
```


Now, just to reorganize the data.
```{r}
hs.df <- hs.df %>% select(subject_id, age, gender, ethnicity, first_careunit, elixhauser, icu_los, 
                          oasis, vasopressor_flag, vent_dur, hyperoxic_dur, 
                          hyperoxic_group, hospital_expire, oneyear_expire)
```

## Exploratory Data Analysis

### Covariates
Lets convert are binary variables to factors.
```{r}
hs.eda.df <- hs.df
hs.eda.df$gender <- as.factor(hs.eda.df$gender)
hs.eda.df$ethnicity <- as.factor(hs.eda.df$ethnicity)
hs.eda.df$first_careunit <- as.factor(hs.eda.df$first_careunit)
hs.eda.df$vasopressor_flag <- as.factor(hs.eda.df$vasopressor_flag)
hs.eda.df$hyperoxic_group <- as.factor(hs.eda.df$hyperoxic_group)
hs.eda.df$hospital_expire <- as.factor(hs.eda.df$hospital_expire)
hs.eda.df$oneyear_expire <- as.factor(hs.eda.df$oneyear_expire)
```


And now lets generate a Table 1.
```{r}
CreateTableOne(vars=c("age","gender", "ethnicity", "first_careunit", "elixhauser", "icu_los", "oasis", "vasopressor_flag","vent_dur", "hyperoxic_dur", "hyperoxic_group"),
               data=hs.eda.df)
```

And because we will be analyzing these patients based on hyperoxic group, lets
see if there are differences between the covariates by group.
```{r}
CreateTableOne(vars=c("age","gender", "ethnicity", "first_careunit", "elixhauser", "icu_los", "oasis",
                      "vasopressor_flag","vent_dur"), strata = "hyperoxic_group",
               data=hs.eda.df)
```

All covariates vary by hyperoxic_group, and clearly will need to be controlled 
for.

As expected, they appear quite related. 

### Other Questions
Before proceeding to examine the outcomes, some useful questions to answer are:

1) How many SpO2 readings are there per patient on average (which we used to
build the durations)?
```{r}
readings.count <- spo2.tidy.df %>% group_by(subject_id) %>% 
  summarise(count = n())
summary(readings.count$count)
```

On average there are ~99 readings with a median of 54 and no less than 4; there
is therefore a good amount of data per patient. And the distributon:

```{r}
ggplot(readings.count) + geom_histogram(aes(x = count))
```


2) What does the distribution of SpO2, hyperoxia duration, ventilation duration
look like?
```{r}
ggplot(spo2.tidy.df) + geom_histogram(aes(x = valuenum), binwidth = 1)
ggplot(hs.eda.df) + geom_histogram(aes(x = hyperoxic_dur))
ggplot(hs.eda.df) + geom_histogram(aes(x = vent_dur))
```

3) Are hyperoxic duration, vent duration, and LOS as related as we would expect?

```{r}
ggplot(hs.eda.df) + geom_point(aes(x= hyperoxic_dur, y = vent_dur,
                                   size = icu_los, alpha = 0.5))
```


### Outcomes

#### Primary outcome 1: Hospital Expiration
Lets now see the proportion of patients in each group who achieved the ouctome.
```{r}
plot_prop_by_level(hs.eda.df,"hyperoxic_group","hospital_expire")
```


Based on this, at least unadjusted, there appears to be a difference; lets look
at crude odds ratio by group.

```{r}
plot_OR_by_level(hs.eda.df, "hyperoxic_group","hospital_expire")
```


There is a dip at hyperoxic_group 2, followed by increases in level 3 and 4.
This is physiologically acceptable as some hyperoxia may be helpful, but at some
point it may become detrimental.

We'll examine this exposure/outcome relationship with model building after
exploring the second primary outcome.

Other factor variables include gender, ethnicity, and care unit.

We can first examine the proportions of the outcome in the exposure groups as it
changes when stratified for the factors.

```{r}
plot_prop_by_level(hs.eda.df,"hyperoxic_group","hospital_expire", 
                   factor.var2 = "gender")
```

```{r}
plot_prop_by_level(hs.eda.df,"hyperoxic_group","hospital_expire", 
                   factor.var2 = "ethnicity")
```

```{r}
plot_prop_by_level(hs.eda.df,"hyperoxic_group","hospital_expire", 
                   factor.var2 = "first_careunit")
```


```{r}
plot_OR_by_level(hs.eda.df, "gender","hospital_expire", 
                 factor.var2 = "hyperoxic_group")
```

```{r}
plot_OR_by_level(hs.eda.df, "ethnicity","hospital_expire", 
                 factor.var2 = "hyperoxic_group")
```

There is a nice linear trend for all groups except Asians and Blacks.

```{r}
plot_OR_by_level(hs.eda.df, "first_careunit","hospital_expire", 
                 factor.var2 = "hyperoxic_group")
```

Rather interestingly, the point estimates are on different sides of the null
by unit in some cases.

#### Primay Outcome 2: One-year Expire

Lets now see the proportion of patients in each group who achieved the ouctome.
```{r}
plot_prop_by_level(hs.eda.df %>% filter(hospital_expire == 0),
                   "hyperoxic_group","oneyear_expire")
```

Based on this, at least unadjusted, there appears to be a difference in group 
3 and 4, although 4 is actually lower than 3. Lets check the OR.

```{r}
plot_OR_by_level(hs.eda.df %>% filter(hospital_expire == 0), 
                 "hyperoxic_group","oneyear_expire")
```

The pattern persists: group 3 is worse than 2 or 4; this makes less sense than
our previous outcome which seemed plausible. This will have to probed further.


## Model Building

### Hopsital Expiration

We'll build a logistic regression model for the outcome of hospital expiration
and the exposure of hyperoxic group. We will adjust for age, gender, length of
ICU stay, total duration of ventilation, highest ICU first day OASIS, elixhauser,
and the presence or absence of vasopressors at anytime during their ICU stays.

These covariates were picked based upon expert opinion and the model wil not be
built in a sequential manner, but rather include all covariates from the start.

The reference group is hyperoxic group 1, the lowest quartile of exposure.

```{r}
hs.he.glm <- glm(hospital_expire ~ hyperoxic_group + age + gender + ethnicity + first_careunit + icu_los +
                   vent_dur + oasis + elixhauser + vasopressor_flag,
                 family = binomial(link = "logit"), data = hs.eda.df)

summary(hs.he.glm)
```

Lets check the ORs and CI for the ORs.
```{r}
exp(hs.he.glm$coefficients)
exp(confint(hs.he.glm))
```

The OR of hospital death for hyperoxic group 2 is 0.74 [0.43 1.25], for group 3 
is 1.62 [1.04, 2.56], and for group 4 is 2.30 [1.34, 3.98] when adjusted for
age, gender, ICU length of stay, total ventilation duration, OASIS, elixhauser,
and vasopressors during ICU course.

Within the conventional hypothesis testing framework utilizing an alpha of 0.05,
the p-values for the test of the coefficients for our three groups as compared
to the reference group (hyperoxic group 1) and adjusted for all covariates are
0.262, 0.0363, and 0.003 respectively. 

Thus, for hyperoxic groups 3 & 4, the data is inconsistent with the state of 
nature described by the null. There is particularly strong evidence against the
null for group 4.

Before continuing, we will check the fit of our model.
```{r}
he.glm.gof <- hoslem_gof(hs.he.glm)
he.glm.gof$p.value
```
The p-value is >0.05, and thus we fail to reject the null: the model fit is 
adequate.


```{r}
pred.hm <- predict(hs.he.glm, type = c("response"))
hs.eda.df$pred_hm <- pred.hm
hs.he.roc <- roc(hospital_expire ~ pred_hm, data = hs.eda.df)
print(hs.he.roc)
```

```{r warning = FALSE, message = FALSE}
roc.plot.df <- data.frame(sens = hs.he.roc$sensitivities, 
                          spec = hs.he.roc$specificities)
roc.plot.df %>% ggplot() + geom_line(aes(x = 1 - spec, y = sens)) +
  geom_abline(intercept = 0, slope = 1, alpha = 0.2) + ggtitle("AUROC = 0.7265")
```

```{r warning = FALSE, message = FALSE}
separationplot(hs.eda.df$pred_hm, 
               as.numeric(hs.eda.df$hospital_expire == 1), shuffle = TRUE, 
               line = TRUE, type = "line", lwd1 = 1, lwd2 = 1, 
               heading = "Separation Plot for Final Hospital Expiration Model")
```

```{r warning = FALSE, message = FALSE}
heatmap.fit(as.numeric(hs.eda.df$hospital_expire == 1), hs.eda.df$pred_hm, color = TRUE)
```

Based on these data, we have a fairly good model and can assume it is properly
specified.

### One year expire

```{r}
hs.oye.glm <- glm(oneyear_expire ~ hyperoxic_group + age + gender + ethnicity + icu_los + first_careunit +
                   vent_dur + oasis + elixhauser + vasopressor_flag,
                 family = binomial(link = "logit"), data = hs.eda.df %>% filter(hospital_expire == 0))
summary(hs.oye.glm)
```

The above result recapitulates what occured earlier; while all groups have an
OR > 1, there appears to be a trend that contradicts our expected relationship.
The third group is highly significant, but for some reason group 4 is not.

And now we can check the fit of this model.
```{r}
oye.glm.gof <- hoslem_gof(hs.oye.glm)
oye.glm.gof$p.value
```

So the model fit is adequate, and the group 3 association is highly significant,
but the conclusion does not follow physiologically. 

## EDA Part 2: Understanding the One-year Result

It is odd that the effect is so high in group 3 but decreases in group 4. We can
examine this in a few ways. First lets make 10 groups instead of 4 and examine 
the mortality trend.

```{r}
hs.eda.df$hyperoxic_quartile <- with(hs.eda.df, cut(hyperoxic_dur, 
                                breaks = quantile(hyperoxic_dur, 
                                                  probs = seq(0,1, by = (1/20)), 
                                                  na.rm=TRUE), 
                                include.lowest = TRUE))


hs.eda.df %>% filter(hospital_expire == 0) %>% group_by(hyperoxic_quartile) %>% 
  summarise(p.oye = mean(as.numeric(oneyear_expire == 1)), count = n()) %>% 
  ggplot() + geom_point(aes(x = hyperoxic_quartile, y = p.oye, size = count)) +
  theme(axis.text = element_text(angle = 90, hjust = 1))
```


```{r}
hs.oye.glm <- glm(oneyear_expire ~ hyperoxic_quartile + age + gender + icu_los +
                   vent_dur + oasis + elixhauser + vasopressor_flag, family = binomial(link = "logit"), data = hs.eda.df %>% filter(hospital_expire == 0))
summary(hs.oye.glm)
```

```{r}
hs.oye.glm <- glm(oneyear_expire ~ rcs(hyperoxic_dur) + age + gender + icu_los + first_careunit + ethnicity +
                   vent_dur + oasis + elixhauser + vasopressor_flag, family = binomial(link = "logit"), data = hs.eda.df %>% filter(hospital_expire == 0))
summary(hs.oye.glm)
```

## Nonlnearity?

Based on what we saw with one-year expiration above, lets examine hospital 
expiration for a non-linear association.

```{r}
hs.eda.df$hyperoxic_quartile <- with(hs.eda.df, cut(hyperoxic_dur, 
                                breaks = quantile(hyperoxic_dur, 
                                                  probs = seq(0,1, by = (1/20)), 
                                                  na.rm=TRUE), 
                                include.lowest = TRUE))


hs.eda.df %>% group_by(hyperoxic_quartile) %>% 
  summarise(p.he = mean(as.numeric(hospital_expire == 1)), count = n()) %>% 
  ggplot() + geom_point(aes(x = hyperoxic_quartile, y = p.he, size = count)) +
  #geom_smooth(aes(x = hyperoxic_quartile, y = (p.he/(1-p.he)), group =1)) +
  theme(axis.text = element_text(angle = 90, hjust = 1))
```

There appears to be a linear relationship, but significant noise.

```{r}
hs.he.glm <- glm(hospital_expire ~ rcs(hyperoxic_dur) + age + gender + ethnicity + first_careunit + icu_los +
                   vent_dur + oasis + elixhauser + vasopressor_flag,
                 family = binomial(link = "logit"), data = hs.eda.df)

summary(hs.he.glm)
```

Thankfully, restricted cubic splines don't provide strong evidence for a 
non-linear relationship.

## Secondary Outcomes

Based on our table one, there was a difference between groups based on race; is 
race associated with hyperoxic time?

Lets begin with an anova.

```{r}
hd.eth.aov <- aov(hs.eda.df$hyperoxic_dur ~ hs.eda.df$ethnicity)
summary(hd.eth.aov)
```

There is a strong difference between the group means; however, we know that 
that hyperoxic duration is very skewed: we can make an argument via CLT, but since
this is a secondary/exploratory outcome lets check it with a more robust method as well. 

```{r}
hd.eth.aov <- kruskal.test(hs.eda.df$hyperoxic_dur ~ hs.eda.df$ethnicity)
print(hd.eth.aov)
```

So race is clearly associated with how much time patients spend hyperoxic, but 
the question then is, could this be confounded by disease severity, comorbidity, 
and etc? Lets build a linear model to adjust for this.

```{r}
hd.eth.lm <- lm(hyperoxic_dur ~ ethnicity + age + gender + first_careunit +
                  icu_los + vent_dur + oasis + elixhauser + vasopressor_flag,
                data = hs.eda.df)
summary(hd.eth.lm)
```

There is a very strong signal for ethnicity, and ~87.85% of the variance is 
explained. However, we should check our model.

```{r}
p <- plot(hd.eth.lm)
```

Unfortunately, based one the residuals, the assumptions of regression do not 
hold. I hypothesize that this is a result of the massive skew in the hyperoxic 
durations. Perhaps we should use median regression.

```{r}
hd.eth.lm <- rq(hyperoxic_dur ~ ethnicity + age + gender + first_careunit +
                  icu_los + vent_dur + oasis + elixhauser + vasopressor_flag,
                tau = 0.5,
                data = hs.eda.df)
summary(hd.eth.lm)
```

Ethnicitity remains highly associated; we lose a great deal of power, but black
patients still have a very strong signal. The point estimate is 3.46226 in this
more robust model. Similary for Asian pateints we see a point estimate of 4.93575
and a very small p-value.

Also noteable is that gender is independenlty associated with hyperoxic duration
with a very strong signal. Woman appear to be recieving more titration. Both of
these associations could be related to the prevalence of disease processes in 
these patients; we aren't excluding any specific diagnoses.
